Evaluating Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission

### 2025-01-16 paper.qmd
reorganize background, move text to methodology, add text on IS2

#### **Background and data selection**

As a result of community feedback and “hack weeks” organized by NSIDC and UW eScience Institute in 2023[@h5cloud2023], NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar that retrieves surface topography of the Earth’s ice sheets, land and oceans [@NEUMANN2019111325].  Because of its complexity, large size and importance for cryospheric studies we targeted the ATL03 data product. The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument. Each ATL03 file contains 1003 geophysical variables in 6 data groups. Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.  Files were accessed using Python tools commonly used by Earth scientists: h5py and Xarray[@Hoyer2017-su].  h5py is a Python wrapper around the HDF5 C API.  xarray^[`h5py` is a dependency of Xarray]  is a widely used Python package for working with n-dimensional data.  We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.

>>>>

Products from NASA's Ice, Cloud, and Land Elevation Satellite 2 (ICESat-2) mission were some of the first NASA datasets to be migrated from on-premises storage at the National Snow and Ice Data Center Distributed Active Archive Center (NSIDC-DAAC) to Amazon Web Services (AWS) Simple Storage Service (S3).  The ICESat-2 observatory carries the Advanced Topographic Laser Altimeter System (ATLAS), a photon-counting lidar altimeter that measures the round-trip travel time of low-energy green laser light to retrieve the heights of ice sheets, sea ice, land surfaces, vegetation, oceans, and clouds [@NEUMANN2019111325].  The geolocated photon height product (ATL03), which is the basic science product from which other higher-level, surface-specific, products are derived, retrieves the latitude, longitude and ellipsoid heights of photons every 70 cm along six ground tracks as the satellite orbits the Earth.  All ICESat-2 data products are in HDF5 format.  The ICESat-2 mission produces # TerraBytes of data each day.  ATL03 files range between # GB and # TB, with an average file size # GB.  Other surface-specific products tend to have smaller file sizes.  Individual file sizes and the volume of data required for most science goals make traditional download-and-analyse workflows time consuming and require local storage.  However, cloud-based workflows, in which ICESat-2 file objects in AWS S3 buckets are accessed from cloud compute instances in the same AWS region as the data, cannot take full advantage of direct download and scaleability offered by these workflows because of the large latency caused by metadata fragmentation and global api lock.

NSIDC started the Cloud Optimized Format Investigation (COFI) project to improve access to HDF5 from the ICESat-2 mission (and other missions) in response to feedback from the ICESat-2 community and the experiences of particpants attending ICESat-2 hack weeks, organized University of Washington eScience Institute.  The initial effort on the COFI project was made at the 2023 hack week by a collection of developers from NASA DAACs, private companies, and scientists and students [@h5cloud2023].  This collaborative effort has continued led by NSIDC with the HDF Group and NASA engineers.

@@

We have targeted the ATL03, geolocated photon height, product because of its complexity, and large size, and because this data is the input to all standard surface-specific data products generated by NASA, as well as the starting point for research and analysis using ICESat-2 products.  ATL03 HDF5 files contain 1003 variables^[In HDF5, variables are called datasets] organized in # groups^[Define groups here].  The most relevant variable in ATL03 are geolocated photon heights from the ICESat-2 ATLAS instrument.  Although our research was focused on this dataset, most of our findings are applicable to any dataset stored in HDF5 and NetCDF4.

We tested access times to original and different configurations of cloud-optimized HDF5 [ATL03 files](https://its-live-data.s3.amazonaws.com/index.html#test-space/cloud-experiments/h5cloud/) stored in AWS S3 buckets in region us-west-2, the region hosting NASA’s Earthdata Cloud archives.  Files were accessed using Python tools commonly used by Earth scientists: h5py and xarray[@Hoyer2017-su].  h5py is a Python wrapper around the HDF5 C API.  xarray^[`h5py` is a dependency of Xarray]  is a widely used Python package for working with n-dimensional data.  We also tested access times using h5coro, a python package optimized for reading HDF5 files from S3 buckets and kerchunk, a tool that creates an efficient lookup table for file chunks to allow performant partial reads of files.
